Enabling inference for context-dependent models of mutation by bounding the propagation of dependency

### 2022-04-25 writeups/context-dependence.tex
submission template

\documentclass[english]{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{babel}
\usepackage[natbib=true,style=authoryear,backend=biber]{biblatex}
\usepackage[hidelinks]{hyperref}
\usepackage{bm}
\usepackage{todonotes}
\usepackage{authblk}
\renewcommand\Affilfont{\itshape\small}

\addbibresource{context-dependence.bib}

\newcommand{\peter}[1]{\todo[color=green!40]{Peter: #1}}
\newcommand{\erick}[1]{\todo[color=purple!40]{Erick: #1}}

@@

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

\title{Enabling inference for context-dependent models of mutation by bounding the propagation of dependency}
\author[2,3,4,5]{Frederick A Matsen IV}
\author[1,*]{Peter L Ralph}

\affil[1]{Institute of Ecology \& Evolution, Department of Biology, University of Oregon}
\affil[2]{Computational Biology Program, Fred Hutchinson Cancer Research Center, United States}
\affil[3]{Department of Genome Sciences, University of Washington}
\affil[4]{Department of Statistics, University of Washington}
\affil[5]{Howard Hughes Medical Institute}
\affil[*]{Corresponding author: plr@uoregon.edu}

@@

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^L \times |\calS|^L$ matrix $G(L)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^L$ jumps to state $y \in \calS^L$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le L$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+j} = u_j \quad &\text{for } 0 \le j < h \\
  y_{i+j} = v_j \quad &\text{for } 0 \le j < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.
\revpoint{2}{3}

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
\revpoint{2}{5}
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^*
    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
    = \max_x \sum_{y \neq x} G(L)_{x,y} .
\]
Note that we do not actually need to construct $G(L)$ to find $\mu^*$:
since $\mu^*$ must only be greater than this maximum value (not necessarily equal to it),
we can take $\mu^*$ to be $L$ multiplied by the maximum ``per-site'' rate.
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to placing a Poisson process of possible
changes at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_{k-1}) = x$.
Then, the new state $X(s_k)$ is chosen by applying a mutation at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
with probability
\[
q_j = \begin{cases}
    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0 \ge 0$.)
If triple $(\mu(j), u(j), v(j))$ is chosen, then
the substring of $x$ matching $u(j)$ that begins at position $i_k$
is replaced with $v(j)$ to produce $X(s_k)$.

@@

Here is a heuristic argument, depicted in Figure~\ref{fig:Tmers}C.
Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
with $\ell = r$.
Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
Furthermore, the sites up to distance $2 \ell$ to the left of $i$
also depend on these mutations.
We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
This suggests that the following approximation should be good:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{4\ell+1,\ell,\ell}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{4\ell+1,\ell,\ell+1}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}
\revpoint{2}{6}

@@

\paragraph{Boundary conditions} \revpoint{2}{1}
Above, we have chosen to disallow changes whose patterns extend beyond the end of the sequence,
%EM This i.e. seems out of place
i.e., the pattern $u$ (and hence, also the replacement $v$) of each transition triple
must fit entirely within the existing sequence.
Alternatively, we might assume periodic boundary conditions,
which are natural for the Ising or TASEP models.
Or, if the sequence we consider is embedded within a larger, unknown sequence,
then we might assume a statistical model for the flanking sequence,
averaging over the rates implied by states chosen from a stationary distribution.
Our choice here of ``empty'' boundary conditions
has the nice property that the total rate of changes on a given sub-segment of sequence
is monotonically increasing in the length of the overhang,
which gives us a simple check that the overhang is long enough
(by seeing whether the total rate of change has converged).
However, the software also allows periodic boundary conditions.

@@

In other applications, reconstruction of an unknown sequence may be of interest.
For instance, training data might be used to estimate parameters of a noisy transmission line,
facilitating the later reconstruction of noisily transmitted signals.
Alternatively, multiple observations of independently transmitted copies of a single sequence
(e.g., down a phylogenetic tree)
might be used to estimate the process, and from this infer the original (ancestral) sequence
(or sample from its posterior distribution).
There is no conceptual barrier to applying these methods to lattices in higher dimensions,
such as noisy images or particle systems modeling other physical processes.
However, computational costs will grow with the size of the local neighborhood used
even faster than in one dimension. \revpoint{2}{2}

@@

\printbibliography

@@

%%%%%%%%%%%%%%%%%
\includereviews

>>>>

\documentclass[english]{Liebert_Author_revised}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{babel}
\usepackage{natbib}
\usepackage{bm}

@@

\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

\title{Enabling inference for context-dependent models of mutation by bounding the propagation of dependency}
\author{
    Frederick A Matsen IV$^{2,3,4,5}$,
    Peter L Ralph$^{1,*}$\\
$^1$Institute of Ecology \& Evolution, Department of Biology, University of Oregon \\
$^2$Computational Biology Program, Fred Hutchinson Cancer Research Center, United States \\
$^3$Department of Genome Sciences, University of Washington \\
$^4$Department of Statistics, University of Washington \\
$^5$Howard Hughes Medical Institute \\
$^*$To whom correspondence should be addressed: plr@uoregon.edu
}

@@

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^L \times |\calS|^L$ matrix $G(L)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^L$ jumps to state $y \in \calS^L$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le L$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+j} = u_j \quad &\text{for } 0 \le j < h \\
  y_{i+j} = v_j \quad &\text{for } 0 \le j < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^*
    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
    = \max_x \sum_{y \neq x} G(L)_{x,y} .
\]
Note that we do not actually need to construct $G(L)$ to find $\mu^*$:
since $\mu^*$ must only be greater than this maximum value (not necessarily equal to it),
we can take $\mu^*$ to be $L$ multiplied by the maximum ``per-site'' rate.
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to placing a Poisson process of possible
changes at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_{k-1}) = x$.
Then, the new state $X(s_k)$ is chosen by applying a mutation at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
with probability
\[
q_j = \begin{cases}
    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0 \ge 0$.)
If triple $(\mu(j), u(j), v(j))$ is chosen, then
the substring of $x$ matching $u(j)$ that begins at position $i_k$
is replaced with $v(j)$ to produce $X(s_k)$.

@@

Here is a heuristic argument, depicted in Figure~\ref{fig:Tmers}C.
Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
with $\ell = r$.
Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
Furthermore, the sites up to distance $2 \ell$ to the left of $i$
also depend on these mutations.
We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
This suggests that the following approximation should be good:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{4\ell+1,\ell,\ell}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{4\ell+1,\ell,\ell+1}(t;\, x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}

@@

\paragraph{Boundary conditions}
Above, we have chosen to disallow changes whose patterns extend beyond the end of the sequence,
%EM This i.e. seems out of place
i.e., the pattern $u$ (and hence, also the replacement $v$) of each transition triple
must fit entirely within the existing sequence.
Alternatively, we might assume periodic boundary conditions,
which are natural for the Ising or TASEP models.
Or, if the sequence we consider is embedded within a larger, unknown sequence,
then we might assume a statistical model for the flanking sequence,
averaging over the rates implied by states chosen from a stationary distribution.
Our choice here of ``empty'' boundary conditions
has the nice property that the total rate of changes on a given sub-segment of sequence
is monotonically increasing in the length of the overhang,
which gives us a simple check that the overhang is long enough
(by seeing whether the total rate of change has converged).
However, the software also allows periodic boundary conditions.

@@

In other applications, reconstruction of an unknown sequence may be of interest.
For instance, training data might be used to estimate parameters of a noisy transmission line,
facilitating the later reconstruction of noisily transmitted signals.
Alternatively, multiple observations of independently transmitted copies of a single sequence
(e.g., down a phylogenetic tree)
might be used to estimate the process, and from this infer the original (ancestral) sequence
(or sample from its posterior distribution).
There is no conceptual barrier to applying these methods to lattices in higher dimensions,
such as noisy images or particle systems modeling other physical processes.
However, computational costs will grow with the size of the local neighborhood used
even faster than in one dimension.

@@

\bibliographystyle{plainnat-revised}
\bibliography{context-dependence}

@@
